---
title: "Candidate Dataset Search: AD vs. Control Prefrontal Cortex RNA-seq"
subtitle: "For expression analysis of a custom noncoding region on chr15"
author: "Your name"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  eval: true
  warning: false
---

## Objective

We're looking for public RNA-seq datasets suitable for comparing expression
of a noncoding region on chromosome 15 (chr15:25,028,445-25,030,175) between
Alzheimer's disease (AD) patients and healthy controls.

**What we're looking for:**

| Criterion | Target |
|---|---|
| Library prep | Random-primed (not poly-A selected); long-read sequencing preferred if available |
| Tissue | Prefrontal cortex (PFC) / DLPFC |
| Groups | AD vs. control, age-matched |
| Sample size | 30-50 samples per group |
| Covariates | Sex (optional) |
| Annotation | This region isn't in standard reference annotations, so it will need custom quantification regardless of which dataset(s) we use |

Each section below covers one search approach: what we did, why, and what
was found. Code is included but collapsed by default -- click "Show code"
under any section to see exactly how a result was produced.

---

## 1. Public repository search: three complementary channels

Rather than relying on one narrow style of search, this section runs
three different, independent approaches and combines whatever each one
finds. Each channel has a different blind spot, so running all three
catches more than any one alone would:

**A. SRA free-text search** -- searches study titles/descriptions for
exact phrase combinations (e.g. "Alzheimer" + "prefrontal cortex"). Fast,
but blind to studies that phrase things differently, or that only mention
tissue in a metadata field this search doesn't read.

**B. GEO index search** -- GEO records include a free-text summary and
"overall design" section that authors write themselves, which is far
richer than SRA's title/description fields. Searching GEO directly (then
resolving matches back to their SRA accessions) catches studies that
never would have matched channel A's stricter phrase search.

**C. Literature search** -- searches PubMed directly for relevant papers,
then pulls the GEO/SRA accession straight out of each paper's own
listed dataset ID. This catches datasets whose paper describes the
tissue/cohort clearly in the abstract, even if the deposited record's own
title or description is terse or non-descriptive (a common issue --
depositors often don't write good metadata).

Every candidate found by any channel goes into one combined list, tagged
with which channel(s) found it -- so nothing gets silently dropped just
because one search style happened to miss it.

```{r}
#| label: setup-reticulate
#| echo: false
# Points R's Python interface at this project's virtual environment.
# Adjust the path if your .venv lives somewhere else.
reticulate::use_python(
  "/Users/heatherkates/Downloads/ad-noncoding-expression/.venv/bin/python",
  required = TRUE
)
```

```{python}
#| label: channel-a-sra-search
import pandas as pd
from pysradb import SRAweb
db = SRAweb()

# Channel A: SRA free-text search, several phrasings
sra_queries = [
    '"Alzheimer" AND "prefrontal cortex"',
    "\"Alzheimer's disease\" AND \"DLPFC\"",
    '"Alzheimer" AND "frontal cortex"',
    '"dementia" AND "prefrontal cortex"',
    '"Alzheimer" AND "BA9"',
    '"Alzheimer" AND "BA10"',
    '"Alzheimer" AND "middle frontal gyrus"',
    '"Alzheimer" AND "superior frontal gyrus"',
]

sra_hits = []
for q in sra_queries:
    r = db.search_sra(search_str=q)
    if r is None or len(r) == 0:
        continue
    r["query_used"] = q
    r["found_via"] = "A: SRA search"
    sra_hits.append(r)

channel_a = pd.concat(sra_hits, ignore_index=True) if sra_hits else pd.DataFrame()
```

```{python}
#| label: channel-b-geo-search
#| eval: false
# PAUSED: this query syntax was wrong -- space-separated terms are read as
# an implicit AND of all six words, an even narrower search than what we
# were trying to fix (0 results). Needs proper boolean OR syntax before
# it's worth re-running; not a priority given Channel C already found
# real candidates below.
from pysradb.search import GeoSearch

geo_search = GeoSearch(
    organism="Homo sapiens",
    geo_query="Alzheimer prefrontal cortex DLPFC frontal cortex RNA-seq",
    geo_dataset_type="expression profiling by high throughput sequencing",
    return_max=200,
)
geo_search.search()
channel_b = geo_search.get_df()
```

```{python}
#| label: channel-c-literature-search
import requests
import time

# Channel C: search PubMed directly, then pull the GEO/SRA accession out
# of each paper's own record.
pubmed_query = (
    '(Alzheimer[Title/Abstract] OR "Alzheimer\'s disease"[Title/Abstract] '
    'OR dementia[Title/Abstract]) AND '
    '(prefrontal[Title/Abstract] OR DLPFC[Title/Abstract] '
    'OR "frontal cortex"[Title/Abstract] OR BA9[Title/Abstract] '
    'OR BA10[Title/Abstract] OR "frontal gyrus"[Title/Abstract]) AND '
    '(RNA-seq[Title/Abstract] OR "RNA sequencing"[Title/Abstract] '
    'OR transcriptom*[Title/Abstract])'
)

esearch = requests.get(
    "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi",
    params={"db": "pubmed", "term": pubmed_query, "retmax": 300, "retmode": "json"},
).json()
pmids = esearch.get("esearchresult", {}).get("idlist", [])

# Pace requests -- NCBI allows ~3/sec without an API key; going faster
# than that is what triggered the 502 errors last run.
lit_hits = []
for pmid in pmids:
    try:
        ids = db.pmid_to_identifiers(pmid)
    except Exception:
        continue
    if ids is not None and len(ids):
        ids["pmid"] = pmid
        ids["found_via"] = "C: Literature search"
        lit_hits.append(ids)
    time.sleep(0.35)

channel_c = pd.concat(lit_hits, ignore_index=True) if lit_hits else pd.DataFrame()

# Save immediately -- don't let a downstream step lose this if it fails.
channel_c.to_csv("candidate_literature_raw.csv", index=False)
```

**For manual review:** rather than running more automated NCBI queries
against each of these (which is what triggered the rate-limiting above),
here's the deduplicated list of unique papers/accessions Channel C found
-- worth a quick manual scan of titles for anything obviously on- or
off-topic before spending more automation effort on it.

```{r}
#| label: channel-c-dedup-table
#| echo: false
if (file.exists("candidate_literature_raw.csv")) {
  lit_raw <- read.csv("candidate_literature_raw.csv")
  id_col <- intersect(c("gse_ids", "srp_ids", "prjna_ids"), names(lit_raw))
  cat(sprintf("**%d** paper-to-accession matches found (before dedup).\n\n", nrow(lit_raw)))
  knitr::kable(lit_raw |> dplyr::distinct(dplyr::across(dplyr::any_of(c(id_col, "pmid")))))
} else {
  cat("_Run the literature search chunk above first._")
}
```

```{r}
#| label: channel-c-download
#| echo: false
#| results: asis
if (file.exists("candidate_literature_raw.csv")) {
  cat(xfun::embed_file("candidate_literature_raw.csv", text = "Download literature-derived candidates (CSV)"))
}
```

```{python}
#| label: filter-channel-a
# Channel A is the only channel feeding the automated filter pipeline
# right now -- Channel B is paused (broken query), and Channel C's results
# are handled separately above as a manual-review list rather than run
# through more automated NCBI calls.
combined = channel_a.copy()

# Human samples only -- catches false positives like mouse AD-model studies
combined_human = combined[combined["organism_name"] == "Homo sapiens"] if len(combined) else combined

# RNA-seq only -- free-text search also pulls in non-RNA-seq assays (e.g.
# a DNA-methylation/Bisulfite-Seq study that happened to mention the right
# tissue and disease terms).
combined_rna = combined_human[
    combined_human["library_strategy"].str.contains("RNA-Seq", case=False, na=False)
] if len(combined_human) else combined_human

# Flag likely single-cell/single-nucleus studies rather than silently
# dropping or including them -- client confirmed bulk RNA-seq is the
# target, so this column makes it a one-glance filter instead of a
# per-title manual read.
sc_pattern = "single.cell|single.nuclei|single.nucleus|snRNA|scRNA|10x genomics|sn-rna|sc-rna"

# Flag specialized RNA-modification/IP assays (e.g. m6A-IP / MeRIP-seq) --
# these report library_strategy = "RNA-Seq" but aren't whole-transcriptome
# expression profiling.
mod_pattern = "m6a|merip|meripseq|rna.ip|immunoprecipitation"

if len(combined_rna):
    combined_rna = combined_rna.copy()
    combined_rna["likely_single_cell"] = (
        combined_rna["study_title"].str.contains(sc_pattern, case=False, na=False, regex=True)
        | combined_rna["experiment_title"].str.contains(sc_pattern, case=False, na=False, regex=True)
    )
    combined_rna["likely_specialized_assay"] = (
        combined_rna["study_title"].str.contains(mod_pattern, case=False, na=False, regex=True)
        | combined_rna["experiment_title"].str.contains(mod_pattern, case=False, na=False, regex=True)
    )

human_studies = combined_rna["study_accession"].unique().tolist() if len(combined_rna) else []

combined_rna.to_csv("candidate_studies_sra.csv", index=False)
```

**Results:**

```{r}
#| label: sra-results-table
#| echo: false
if (file.exists("candidate_studies_sra.csv")) {
  sra_results <- read.csv("candidate_studies_sra.csv")
  cat(sprintf("**%d** human RNA-seq runs found, across **%d** distinct studies.\n\n",
              nrow(sra_results), length(unique(sra_results$study_accession))))
  knitr::kable(
    sra_results |>
      dplyr::distinct(study_accession, study_title) ,
    col.names = c("Study accession", "Study title")
  )
} else {
  cat("_No results yet -- run the search above first._")
}
```

```{r}
#| label: sra-download
#| echo: false
#| results: asis
if (file.exists("candidate_studies_sra.csv")) {
  cat(xfun::embed_file("candidate_studies_sra.csv", text = "Download SRA candidate list (CSV)"))
}
```

---

## 2. Consortium search: AMP-AD / AD Knowledge Portal (Synapse)

**What we did:** Logged into Synapse and located the AMP-AD RNAseq
Harmonization Study, which combines RNA-seq data from the three major
Alzheimer's brain-tissue cohorts (ROSMAP, MSBB, and Mayo Clinic) into one
uniformly processed dataset.

**Why:** This is the strongest candidate source for the 30-50 samples per
group target -- these cohorts include hundreds of prefrontal-cortex samples
each, with matched age and sex information already collected. Public SRA
search (Section 1) is unlikely to surface a study this size on its own,
since access to these cohorts requires a data use request rather than a
plain download.

```{r}
#| label: synapse-query
#| eval: false
# eval: false until Synapse login + data access approval are in place --
# flip to true once that's set up.
synLogin()

# AMP-AD RNAseq Harmonization Study (ROSMAP + MSBB + Mayo combined)
harmonized_study <- "syn21241740"

children <- as.list(synGetChildren(harmonized_study))
children_df <- do.call(rbind, lapply(children, function(x) {
  data.frame(name = x$name, id = x$id, type = x$type, stringsAsFactors = FALSE)
}))

metadata_table_id <- "REPLACE_WITH_TABLE_ID_FROM_ABOVE"
```

```{r}
#| label: synapse-metadata-pull
#| eval: false
meta <- synTableQuery(
  sprintf("SELECT * FROM %s WHERE tissue LIKE '%%prefrontal%%'", metadata_table_id)
)$asDataFrame()

pfc_ad_control <- meta |>
  dplyr::filter(diagnosis %in% c("AD", "Control")) |>
  dplyr::filter(!is.na(age_death), !is.na(sex))

write.csv(pfc_ad_control, "candidate_samples_synapse.csv", row.names = FALSE)
```

**Status:** In progress -- Synapse access is authenticated, but data-use
approval for the controlled-access cohorts (ROSMAP, MSBB, Mayo) is still
being finalized, and the specific metadata table ID needs to be confirmed.
Results will populate here automatically once that's complete.

```{r}
#| label: synapse-download
#| echo: false
#| results: asis
if (file.exists("candidate_samples_synapse.csv")) {
  cat(xfun::embed_file("candidate_samples_synapse.csv", text = "Download AMP-AD candidate samples (CSV)"))
} else {
  cat("_Download link will appear here once Synapse data access is complete._")
}
```

---

## 3. Candidate dataset summary

This table is the running answer to "which datasets fit the brief" --
updated as each search source above completes.

```{r}
#| label: summary-table
#| echo: false
summary_tbl <- tibble::tribble(
  ~Dataset,       ~Source,          ~Tissue,     ~`AD (n)`, ~`Control (n)`, ~`Library prep`,      ~`Long-read`, ~Status,
  "ROSMAP",       "AMP-AD/Synapse", "DLPFC",      NA,         NA,            "Poly-A (typical) - confirming", "No",  "Pending data access",
  "MSBB",         "AMP-AD/Synapse", "PFC (BM10)", NA,         NA,            "Poly-A (typical) - confirming", "No",  "Pending data access",
  "Mayo RNAseq",  "AMP-AD/Synapse", "TCX / CER",  NA,         NA,            "Poly-A (typical)",   "No",  "Likely excluded - wrong tissue region",
)
knitr::kable(summary_tbl)
write.csv(summary_tbl, "candidate_dataset_summary.csv", row.names = FALSE)
```

```{r}
#| label: summary-download
#| echo: false
#| results: asis
cat(xfun::embed_file("candidate_dataset_summary.csv", text = "Download dataset summary (CSV)"))
```

---

## 4. How we'll measure the target region

The chr15 region of interest isn't in standard gene annotations
(GENCODE/RefSeq), so once candidate datasets are selected, expression will
be quantified directly from raw sequencing coverage at those exact
coordinates rather than relying on a pre-built gene model. For any dataset
already processed through the public `recount3` resource, this can be done
without re-downloading or re-aligning raw sequencing files.

```{r}
#| label: recount3-setup
#| eval: false
saf <- read.delim("Target_chr15_sites.saf")
target_gr <- GenomicRanges::GRanges(
  seqnames = saf$Chr,
  ranges = IRanges::IRanges(start = saf$Start, end = saf$End),
  strand = saf$Strand
)
```

---

## 5. Important considerations

- **Library prep:** Most AMP-AD RNA-seq is poly-A selected rather than
  random-primed. We're confirming per-sample library prep before assuming
  full coverage of the target region -- if it turns out coverage is
  incomplete, that would narrow which cohorts are usable.
- **Long-read sequencing:** No public long-read PFC dataset currently
  exists at the 30-50 samples/group scale. The best available option (a
  12-sample Oxford Nanopore study) is useful only as a secondary check on
  isoform structure, not as a primary analysis cohort.
- **Data access timeline:** AMP-AD cohorts are controlled-access and
  require a data use request through Synapse, which can take some time to
  clear -- this is the main open item holding up a final sample-size count.
